Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-05 documents/ms.tex
a few comments

The uncertainties produced by the GP method are more representative than those
produced by the ACF and periodogram methods because the posterior PDF of the
period is explored using MCMC.
Whilst they are an improvement, unfortunately these uncertainties are still
underestimated in the majority cases.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lie within 1 $\sigma$ of the true period.
50\% lie within 2 $\sigma$ and 66\% within 3 $\sigma$.
The largest outlier is 114 $\sigma$ away from the true value.
In the majority of these cases the uncertainties are underestimated due to the
multi-modal nature of the period posterior PDFs, which makes them difficult to
sample.
The underestimated uncertainties can also be attributed to the model choice.
Although a GP model is more appropriate than a sinusoid, it is still an
imperfect, effective model.
A perfect physical model of the star would still produce better results with
more representative uncertainties.

@@

We attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method, a Lomb-Scargle
periodogram method and an autocorrelation function method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the standard version of the ACF method, most commonly
implemented in the literature, often slightly under-predicts the rotation
period due to a subtle feature of its peak detection algorithm.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2014}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and LS periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
approximate because in many cases the posterior PDFs of the parameters are
multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, and a more accurate
uncertainty would result.
Unfortunately, due to realistic computational resources and the limitations of
MCMC sampling techniques for multi-modal posteriors, we are not always able to
map out the entire posterior PDFs.
In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF of the periodic component of the
covariance matrix, not the actual rotation period of a physical star.
We find that only a quarter of the reported uncertainties are accurate.

@@

% acknowledgements
This research was funded by the Simons Foundation and the Leverhulme Trust.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
Kepler mission is provided by the NASA Science Mission directorate.

>>>>

The uncertainties produced by the GP method are more representative than those
produced by the ACF and periodogram methods because the posterior PDF of the
period is explored using MCMC.
Whilst they are an improvement, unfortunately these uncertainties are still
underestimated in the majority cases.
Of the rotation periods recovered from the simulated light curves, only 25\%
of the measured periods lie within 1 $\sigma$ of the true period.
50\% lie within 2 $\sigma$ and 66\% within 3 $\sigma$.
\tdmcomment{What are these numbers for the uninformed-prior experiment?}
The largest outlier is 114 $\sigma$ away from the true value.
In the majority of these cases the uncertainties are underestimated due to the
multi-modal nature of the period posterior PDFs, which makes them difficult to
sample. \tdmcomment{Is there any evidence of this?  Doesn't sound convincing to me.}
The underestimated uncertainties can also be attributed to the model choice.
Although a GP model is more appropriate than a sinusoid, it is still an
imperfect, effective model.
A perfect physical model of the star would still produce better results with
more representative uncertainties.

@@

We attempted to recover the rotation periods of \naigrain\ simulated
\kepler-like light curves for solid-body rotators \citep{Aigrain2015} using
three different methods: a new Gaussian process method, a Lomb-Scargle
periodogram method and an autocorrelation function method.
We demonstrate that the GP method produces the most accurate rotation periods
of the three techniques, providing a large improvement over the LS periodogram
method and a moderate one over the ACF method.
We also find that the standard version of the ACF method, most commonly
implemented in the literature, often slightly under-predicts the rotation
period due to a subtle feature of its peak detection algorithm.
In addition, we compared the rotation periods of \kepler\ objects of
interest measured using the GP method to those measured previously by
\citet{Mcquillan2013}.
The good agreement between the two sets of results demonstrates that the GP
method works well on real \kepler\ data.

Unlike the ACF and LS periodogram methods, the GP method provides posterior
PDF samples which can be used to estimate rotation periods uncertainties.
Although an improvement on competing methods, these uncertainties are still
approximate because in many cases the posterior PDFs of the parameters are
multi-modal.
If it were possible to run an MCMC sampler for an infinite number of steps,
the entire posterior PDF would eventually be sampled, and a more accurate
uncertainty would result.
Unfortunately, due to realistic computational resources and the limitations of
MCMC sampling techniques for multi-modal posteriors, we are not always able to
map out the entire posterior PDFs.
\tdmcomment{I don't think this is the problem;
I think the fits that we judge to have converged have actually converged.}

In addition, although the GP model used here is clearly a good one, it is
still only an effective model, not an accurate physical model.
It can only capture the posterior PDF of the periodic component of the
covariance matrix, not the actual rotation period of a physical star.
We find that only a quarter of the reported uncertainties are accurate.

@@

% acknowledgements
This research was funded by the Simons Foundation and the Leverhulme Trust.
TDM is supported by NASA grant NNX14AE11G, and acknowledges the 
hospitality of the Institute for Advanced Study, 
where part of this work was completed.
Some of the data presented in this paper were obtained from the Mikulski
Archive for Space Telescopes (MAST).
STScI is operated by the Association of Universities for Research in
Astronomy, Inc., under NASA contract NAS5-26555.
Support for MAST for non-HST data is provided by the NASA Office of Space
Science via grant NNX09AF08G and by other grants and contracts.
This paper includes data collected by the Kepler mission. Funding for the
Kepler mission is provided by the NASA Science Mission directorate.
